Dish: Distribution-oblivious Shell Scripting

### 2019-11-22 pldi/draft.tex
Write extensibility subsection

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}
\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
\newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
\newcommand{\review}[1]{{\color{red}#1}}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
\caption{
  \textbf{\sys transformation overview.}
	\sys XYZ
}
\label{fig:overview}
\end{figure}

@@

\section{Extensibility}
\label{ext}

Here is the type for \ttt{bwa}, a command that performs a Burrows-Wheeler transform over genomic data:
% ( bwa ) :: ( Pure )  => { In \/ File* } -> { Out /\ Err }    -- default case
%    | -h :: ( Sless ) => { Out }                              -- if needed
\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
(bwa) :: (Pure) ($$\Rightarrow$$){In($$\lor$$)File*}($$\rightarrow$$){Out($$\land$$)Err}
  |-h :: (Sless)($$\Rightarrow$$){Out}                              
\end{lstlisting}

It states that the command defaults to the Pure class.
It's operation can be thought as a transformation from input streams to output streams.
It operates either on stdin or, if this doesn't exist, one or more files specified as arguments;
% TODO: check actual man page for how to specify files
  it writes to the output and error stream rather than a file.
The \ttt{-h} flag moves \ttt{bwa} into the Stateless class;
  its output is a constant function.

ULTRA DRAFTY:

\kk{Text point: A problem that is specific to the shell is
  extensibility.}

The major characteristic that makes the shell popular and very useful
is its extensibility, as also mentioned in \Cref{bg}. Commands change,
get extended, and each user continuously installs more commands to use
in their system.

\kk{Text point: This restricts the space of solutions for distribution
  of shell scripts}

This means that any one-time solution is not satisfactory, since it
would very easily become obsolete, or it would only handle a limited
amount of shell scripts.

\kk{Text point: The characteristics of a good solution}

A solution that applies to shell programms should thus be extensible,
in the sense that adding or changing a command, doesn't invalidate the
assumptions that the tool makes. In our case, the information about
the command categories. A developer of a new command, should be able
to categorize it accordingly, and succinctly describe how the system
can distribute a command.

\kk{Solution: A lightweight language that can be used by a command
  developer to categorize it for distribution.}

We address this issue by designing a lightweight language whose
purpose is to categorize a command, describes how it reads its input
(meaning in what order), and how to distribute it (in case it is
pure).

%% \kk{I am not sure if the order in which a command reads its
%%   input will be part of the language since many commands can be made
%%   to read from one place (like stdin or their first argument). Even in
%%   that case though, where the command only reads from one argument, we
%%   have to make sure that there is a way for the user to pinpoint which
%%   argument does a command read from.}

\subsection{Categorization Language}

\kk{Text point: Introduce problem: Most shell commands, belong to
  different classes depending on their flags, arguments}

Consider the command \ttt{cat} \kk{Maybe that is not he best
  example. TODO: Find a good example of a simple command that is in
  different classes depending on the arguments.}. When used without
any flag it is stateless, etc...

\kk{Text point: Our language must be able to categorize commands to
  different classes according to their arguments.}

\kk{We should have a crisp point about why we have this categorization
  language, and why don't we just allow someone to write a function in
  python for each command, that given the command and its arguments,
  returns whether the command category. Possible arguments include,
  the fact that this language is simpler to use, especially by non
  experts that just run some script but have installed some commands
  that are not ``supported''. Another possible argument is that we
  could be able to reason about the constructs of the categorization
  lagnauge, and that they will be easier to read and
  understand. Another argument is that since these categorizations
  should be shareable among users, it would be bad to execute
  arbitrary python code, so using this language, expressivity is
  limited. All of these arguments are a little bit weak though. Niko,
  what do you think?}

\kk{New sketch of the language}

We assume that commands satisfy the standard that options and flags
come first, and then file arguments. The language contains a record
for each command name, together with a sequence of predicates that map
to command categories and designation of input output.

A record contains a sequence of predicates => categories. Each
predicate is a combination of not, or, and and the atomic predicates
are the presence of a flag, or an equality that checks that the value
of a flag is equal to something (because flags can also have
values). The final predicate can be left empty, and it is trivially true.

Pythonic range for several inputs/outputs

@@

\begin{grammar}
  <flag> ::= `-' <string>

@@

  <arg> ::= `arg[' <int> `]'

  <input> ::= `stdin' | <arg>

@@

  <flag-pred> ::= <flag>
  \alt `value' <flag> = <string>
  \alt `!' <flag-pred>
  \alt <flag-pred> `or' <flag-pred>
  \alt <flag-pred> `and' <flag-pred>

  <assignment> ::= `(' <category>, <input>, <output> `)'

  <predicate> ::= <flag-pred> `=>' <assignment>

@@

  <command> ::= <name> `:' <pred-list>

\end{grammar}

\kk{We don't need to designate the stderror, because we assume that it
  is never the main output of a command and it will never be used by
  the input of another command in the pipeline. If someone indeed
  wants to do this, they can just use a redirect I think to get around
  it.}

\kk{Give an example of a flag that has value and how this could alter
  category or input etc...}

Interpreter:

Given a command and the set of its arguments, the intepreter returns
the command category. In does so by returning the category of the first
predicate that matches.

\kk{If we talk about the language it would be good to give some
  statistics about how many commands (out of the ones that we have
  classified) can be represented with one, two clauses etc. If we have
  a different solution (like python function that given a command and
  its arguments returns the category) then we could talk about how
  many lines of code it took to categorize all the commands that we
  did.}

As described in distributability\ref{}, the abstract notion of a
program that we have is programs with one input and one output
stream. The command developer must be able to express the input
argument of a commmand (or if it is the stdin).

\kk{Note: Since a categorization language will probably not be
  complete (especially the one showing the input arguments) there
  should be a backup mechanism, when the language is not expressive
  enough, like a python function that categorizes the command, and
  returns its input argument if it is stateless.}

\subsection{Pure Commands}

@@

\kk{Can we find a solution for the commands in coreutils?}

\kk{Important TODO: Find at least one more command that can be
  executed like sort. Does wc work? If there is no pure command that
  can be easily parallelized other than sort, this story is not very
  nice.}

@@

>>>>

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}
\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
\newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
\newcommand{\review}[1]{{\color{red}#1}}
\newcommand{\tr}[1]{} %% Text and comments for technical report

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
\caption{
  \textbf{\sys transformation overview.}
	\sys XYZ \todo{This is slightly cropped?}
}
\label{fig:overview}
\end{figure}

@@

\subsection{Extensibility}
\label{ext}

%% Here is the type for \ttt{bwa}, a command that performs a Burrows-Wheeler transform over genomic data:
%% % ( bwa ) :: ( Pure )  => { In \/ File* } -> { Out /\ Err }    -- default case
%% %    | -h :: ( Sless ) => { Out }                              -- if needed
%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%% (bwa) :: (Pure) ($$\Rightarrow$$){In($$\lor$$)File*}($$\rightarrow$$){Out($$\land$$)Err}
%%   |-h :: (Sless)($$\Rightarrow$$){Out}                              
%% \end{lstlisting}

%% It states that the command defaults to the Pure class.
%% It's operation can be thought as a transformation from input streams to output streams.
%% It operates either on stdin or, if this doesn't exist, one or more files specified as arguments;
%% % TODO: check actual man page for how to specify files
%%   it writes to the output and error stream rather than a file.
%% The \ttt{-h} flag moves \ttt{bwa} into the Stateless class;
%%   its output is a constant function.

One of the major characteristics of the shell is its extensibility, as
also mentioned in \Cref{bg}. Existing commands are modified and
extended over time and users are free to install custom commands from
a wide variety of sources.
%% \kk{Text point: This restricts the space of solutions for
%%   distribution of shell scripts}
This means that any one-time solution for categorizing shell commands
is not satisfactory, since it would very quickly become obsolete,
being able to support a very limited amount of shell scripts.

%% \kk{Text point: The characteristics of a good solution}

A solution that applies to shell programms should thus be extensible,
in the sense that adding or changing a command, doesn't invalidate the
existing information on command categories. A developer of a new
command, should be able to categorize it accordingly, and succinctly
express the information that the system needs in order to distribute
it.

%% \kk{Solution: A lightweight language that can be used by a command
%%   developer to categorize it for distribution.}

\TODO{Make sure that before this point, we mention that our notion of
  command has one input and one output. Concretely: the abstract
  notion of a program that we have is programs with one input and one
  output stream. The command developer must be able to express the
  input argument of a commmand (or if it is the stdin).  }

We address this issue by designing a lightweight language whose
purpose is to categorize a command and describe the sequence of
sources from which it reads its input.

%% \subsection{Categorization Language}

%% \kk{Text point: Introduce problem: Most shell commands, belong to
%%   different classes depending on their flags, arguments}

\TODO{Mention the point of these two paragraphs in the intro?}

Consider the command \ttt{comm}. It takes as input two files and
implements some form of join. When invoked without any option, it
produces three-column output. The first column contains lines that
unique to the first file, the second column contains lines that are
unique to the second file, and the last column contains lines that
exist in both files. In the general case \ttt{comm} is pure, since it
needs to wait until it reads both files completely before it can
output its results.

However, \ttt{comm} can also be invoked using some options that
supress the output of any combination of columns. More precisely, the
option \ttt{-1} suppresses the first column of the output. Similarly,
\ttt{-2} and \ttt{-3} supress the second and third column. Using both
\ttt{-2} (or \ttt{-1}) and \ttt{-3}, the command becomes stateless
regarding its first (resp. the second) file argument.

In order to address this, we have designed a categorization language
that is flexible enough to identify commands to different categories
given their options. The complete language syntax can be seen in
\Cref{fig:dsl}.

@@

\begin{figure}
  \centering
  \begin{grammar}
    <option> ::= `-' <string>

@@

    <maybe-int> ::= | <int>

    <arg> ::= `args[' <int> `]'

    <args> ::= <arg>
    \alt `args[' <maybe-int> `:' <maybe-int> `]'

    <input> ::= `stdin' | <args>

    <inputs> ::= <input>
    \alt <input> `,' <inputs>

@@

    <option-pred> ::= <option>
    \alt `value' <option> = <string>
    \alt `not' <option-pred>
    \alt <option-pred> `or' <option-pred>
    \alt <option-pred> `and' <option-pred>

    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'

    <predicate> ::= <option-pred> `=>' <assignment>

@@

    <command> ::= <name> `\{' <pred-list> `\}'

    <command-list> ::= <command>
    \alt <command> <command-list>
  \end{grammar}
  \caption{\todo{Categorization language}}
  \label{fig:dsl}
\end{figure}

We assume that the arguments of a command can always be separated to
options and file arguments (checking if they start with a
\ttt{-}~\footnote{Note that the single character \ttt{-} is not an
  option as it is most often used to designate that a command reads
  from its standard input}). A file written in the language contains a
record containing a mapping from predicates to assignments for each
command. Predicates can be simple boolean formulas on the existence
and the value of all options. Each assignment contains the command
category, the sequence of its input sources, and its output
source. Inputs to commands are a sequence of file arguments and its
standard input. Similarly, the output is either its standard output or
a file argument. An example record for comm can be seen below:

\begin{lstlisting}[float=h, numbers=none]
  comm {
    | -1 and -3 =>
      (stateless, [args[1]], stdout)
    | -2 and -3 =>
      (stateless, [args[0]], stdout)
    | otherwise =>
      (pure, [args[0]], stdout)
  }
\end{lstlisting}

The semantics of the language is straightforward: given a command, the
interpreter collects its options and then returns the assignment of
the first predicate that is satisfied. The final predicate is always
satisfied. For example the following two invocations of comm would be
classified as follows:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 comm -13 f1 f2 # => (stateless, f2, stdout)
 comm f1 f2     # => (pure, f1, stdout)
\end{lstlisting}

\tr{We don't need to designate the stderror, because we assume that it
  is never the main output of a command and it will never be used by
  the input of another command in the pipeline. If someone indeed
  wants to do this, they can just use a redirect I think to get around
  it.}

\tr{If we talk about the language it would be good to give some
  statistics about how many commands (out of the ones that we have
  classified) can be represented with one, two clauses etc. If we have
  a different solution (like python function that given a command and
  its arguments returns the category) then we could talk about how
  many lines of code it took to categorize all the commands that we
  did.}

\tr{Note: Since a categorization language will probably not be
  complete (especially the one showing the input arguments) there
  should be a backup mechanism, when the language is not expressive
  enough, like a python function that categorizes the command, and
  returns its input argument if it is stateless.}

\tr{(Maybe) We should have a crisp point about why we have this categorization
  language, and why don't we just allow someone to write a function in
  python for each command, that given the command and its arguments,
  returns whether the command category. Possible arguments include,
  the fact that this language is simpler to use, especially by non
  experts that just run some script but have installed some commands
  that are not ``supported''. Another possible argument is that we
  could be able to reason about the constructs of the categorization
  lagnauge, and that they will be easier to read and
  understand. Another argument is that since these categorizations
  should be shareable among users, it would be bad to execute
  arbitrary python code, so using this language, expressivity is
  limited. All of these arguments are a little bit weak though. Niko,
  what do you think?}

\paragraph{Pure Commands}

\TODO{Complete this, giving example of sort and wc}

@@

\tr{Can we find a solution for the commands in coreutils?}

@@

\TODO{We have to also cite the paper by Konstantinos, Caleb, Rajeev,
  etc since they also try to do semantics preserving optimizations on
  dataflow graphs}
